Distributional Trees and Forests for Circular Data

### 2019-07-04 pkg/disttree/inst/IWSM2019/slides.Rnw
small improvementes (adapted appearance order in tree theory slide, selected different example plot)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2639 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4687e5057b1e0198dc8d928d1985800d9811b222
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2640 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c3d79d7112537c5b373926c3178ebfc81431adb1

# setwd("~/svn/partykit/pkg/disttree/inst/IWSM2019")
library("disttree")
#library("Formula")
library("latex2exp")
library("circmax")
@

@@

\textbf{Application:} Probabilistic forecasting of wind directions for the next hour at Innsbruck airport (Austria, \textcolor{gray}{\url{https://innsbruck-airport.panomax.com}}).

@@

\begin{frame}[fragile]
\frametitle{Von Mises distribution}
\begin{minipage}{0.5\textwidth}
``Circular normal distribution'' with
\begin{itemize}
\item location parameter $\mu \in [0, 2\pi)$ 
\item concentration parameter $\kappa > 0$ 
\end{itemize}
\end{minipage}
\hspace{1cm}
\begin{minipage}{0.4\textwidth}
%\begin{center}
%\vspace*{-1cm}
\setkeys{Gin}{width=0.6\linewidth}
\includegraphics{density_right.pdf}
%\end{center}
\end{minipage}

@@

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\text{vM}(Y; \theta)$: \\ % to the whole data set:\\
Estimate $\hat{\theta}$ via maximum likelihood\\
$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.23\textwidth}
%\vspace{-0.1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\text{vM}(Y;\hat{\theta}$)};
}
\visible<5->{
\node[inner sep=0pt] (density_root) at (0.8, 1.7)
    {\includegraphics[width=0.6\textwidth]{density_root.pdf}};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2);
}
\visible<9-10>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<11>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$\text{vM}(Y_1;\hat{\theta}_1$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$\text{vM}(Y_2;\hat{\theta}_2$)};
}
\visible<11->{
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<12->{
\node[inner sep=0pt] (density2) at (-0.1,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_left.pdf}};
}
\visible<12->{
\node[inner sep=0pt] (density3) at (1.7,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_right.pdf}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
%\begin{adjustwidth}{-0.0em}{-1em}
\begin{enumerate}
\setcounter{enumi}{2}
\item<8-> Split along the covariate $X$ with strongest association or instability and\\ at breakpoint $p$ with highest improvement in log-likelihood.
\item<10-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}
%\end{adjustwidth}
\vspace{0.3cm}
\visible<13->{
{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
}
\end{frame}

@@

\textbf{Applied models and reference models:}
\begin{itemize}
\item Circular distributional tree
\item Circular distributional forest
\item Persistence
\item Climatology 
\end{itemize}

@@

\textbf{Evaluation setting:} 
\begin{itemize}
\item Data set of $40\ 000$ observations with 354 covariates
\item 5-fold cross validation
\item Results are compared based on circular CRPS (Grimit et al., 2006)
\end{itemize}

@@

\begin{frame}
\frametitle{Software}
%\vspace{0.4cm}
\textbf{Packages:} \emph{disttree} %and \emph{circmax} 
available on R-Forge at\\

@@

Grimit E P, Gneiting T, Berrocal V J, Johnson N A (2006).
  \dquote{The Continuous Ranked Probability Score for Circular Variables and its Application to Mesoscale Forecast Ensemble Verification.}
  \textit{Q.J.R. Meteorol. Soc.}, 
  \textbf{132}, 2925--2942. 
  \doi{10.1256/qj.05.235}

>>>>

# setwd("~/svn/partykit/pkg/disttree/inst/IWSM2019")
library("disttree")
#library("Formula")
library("latex2exp")
#library("circmax")
@

@@

\textbf{Application:} Probabilistic forecasting of wind directions for the next hour at Innsbruck airport (Austria, \textcolor{gray}{\url{https://innsbruck-airport.panomax.com/}}).

@@

\begin{frame}[fragile]
\frametitle{Von Mises distribution}
\begin{minipage}{0.5\textwidth}
``Circular normal distribution'' with
\begin{itemize}
\item location parameter $\mu \in [0, 2\pi)$ 
\item concentration parameter $\kappa > 0$ 
\end{itemize}
\end{minipage}
\hspace{1cm}
\begin{minipage}{0.4\textwidth}
%\begin{center}
%\vspace*{-1cm}
\setkeys{Gin}{width=0.6\linewidth}
\includegraphics{density_left.pdf}
%\end{center}
\end{minipage}

@@

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\text{vM}(Y; \theta)$: \\ % to the whole data set:\\
Estimate $\hat{\theta}$ via maximum likelihood\\
$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.23\textwidth}
%\vspace{-0.1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\text{vM}(Y;\hat{\theta}$)};
}
\visible<5->{
\node[inner sep=0pt] (density_root) at (0.8, 1.7)
    {\includegraphics[width=0.6\textwidth]{density_root.pdf}};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2);
}
\visible<9>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<10>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$\text{vM}(Y_1;\hat{\theta}_1$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$\text{vM}(Y_2;\hat{\theta}_2$)};
}
\visible<10->{
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<11->{
\node[inner sep=0pt] (density2) at (-0.1,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_left.pdf}};
}
\visible<11->{
\node[inner sep=0pt] (density3) at (1.7,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_right.pdf}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
%\begin{adjustwidth}{-0.0em}{-1em}
\begin{enumerate}
\setcounter{enumi}{2}
\item<8-> Split along the covariate $X$ with strongest association or instability and\\ at breakpoint $p$ with highest improvement in log-likelihood.
\item<12-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}
%\end{adjustwidth}
\vspace{0.3cm}
\visible<13->{
{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
}
\end{frame}

@@

\textbf{Novel vs.\ reference models:}
\begin{itemize}
\item Circular distributional tree
\item Circular distributional forest
\item Persistence
\item Climatology 
\end{itemize}

@@

\textbf{Evaluation setting:} 
\begin{itemize}
\item Data set of $40\ 000$ observations with 354 covariates
\item 5-fold cross validation
\item Results are compared based on circular CRPS
\end{itemize}

@@

\begin{frame}
\frametitle{Software}
%\vspace{0.4cm}
\textbf{Software:} \emph{disttree} %and \emph{circmax} 
available on R-Forge at\\

@@

Grimit E P, Gneiting T, Berrocal V J, Johnson N A (2006).
  \dquote{The Continuous Ranked Probability Score for Circular Variables and its Application to Mesoscale Forecast Ensemble Verification.}
  \textit{Quarterly Journal of the Royal Meteorological Society}, 
  \textbf{132}, 2925--2942. 
  \doi{10.1256/qj.05.235}
